HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-09-25 docs/HLL_exp_bad/main.tex
tweakery

%\IEEEtitleabstractindextext{%
\begin{abstract}
Computing the count of distinct elements in massive data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Flajolet \emph{et al.}, 2007) estimates a data set's cardinality while using significantly less memory than a naive approach at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. In this cautionary note, we show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm and has some flexibility in choosing what inputs will be recorded, it can manipulate the cardinality estimate to be exponentially smaller than the true cardinality.
\end{abstract}
%}

@@

\begin{figure}[ht]
\centering
     \makebox[\linewidth]{
\fbox{
\begin{minipage}{40em}
\begin{flushleft}
\textit{Let $h : \mathcal{D} \rightarrow \{0, 1\}^{32}$ hash data from some domain $\mathcal{D}$ to binary $32$-bit strings.\\
Let $\rho(s)$ be the position of the leftmost $1$-bit of $s$: e.g., $\rho(1...) = 1, \rho(0001...) = 4, \rho(0^K) = K + 1.$\\}
\textbf{define} $m = 2^n$ with $n \in [4..16].$\\
\textbf{define} $\alpha_{16} = 0.673$; $\alpha_{32} = 0.697$; $\alpha_{64} = 0.709$; $\alpha_{m} = 0.7213/(1+ 1.079/m)$ for $m \ge 128$;\\
\vspace{0.5em}
\textbf{Algorithm} HyperLogLog (\textbf{input} $\mathcal{M}$ : multiset of items from domain $\mathcal{D}$).\\
\textbf{initialize} a collection of $m$ registers, $M[1], ..., M[m]$, to $0$;\\
\vspace{0.5em}
\textbf{for} $v \in \mathcal{M}$ \textbf{do}\\
    \hspace{2em} \textbf{set} $x := h(v);$\\
    \hspace{2em} \textbf{set} $j = 1 + <x_1x_2...x_{n}>_2$; \hfill \textit{\{the binary address determined by the first $n$ bits of $x$\}}\\
    \hspace{2em} \textbf{set} $w := x_{n+1}x_{n+2}...$;\\
   \hspace{2em} \textbf{set} $M[j] :=$ max$(M[j], \rho(w))$;\\
\vspace{0.2em}
\textbf{compute} $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$; \hfill \textit{\{the raw HyperLogLog estimate\}}\\
\vspace{0.2em}
\textbf{if} $E \le \frac{5}{2}m$ \textbf{then}\\
    \hspace{2em}  \textbf{let} $V$ be the number of registers equal to $0$;\\
    \hspace{2em}  \textbf{if} $V \neq 0$ \textbf{then set} $E^*:=  m\cdot \ln(m/V)$ \textbf{else set} $E^*:= E$; \hfill \textit{\{small range correction\}}\\
\textbf{if} $E \le \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em} \textbf{set} $E^*:= E$; \hfill \textit{\{intermediate range—no correction\}}\\
\textbf{if} $E > \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em}  \textbf{set} $E^*:= -2^{32}$ln$(1 - E/2^{32})$; \hfill \textit{\{large range correction\}}\\
\textbf{return} \textit{cardinality estimate $E^*$ with typical relative error} $\pm1.04/\sqrt{m}$.
\end{flushleft}
\end{minipage}
}}
\caption{The HyperLogLog Algorithm from~\cite{hll}}
\label{fig:hll}
\end{figure}

@@

Here the attack is somewhat more complex. We begin by explaining the core idea, then a simplistic version of our attack, then a more complex but performant version, and finally, we give a heuristic performance analysis of the attack.

@@

\paragraph{Simplistic Attack:}
The simplistic version of our attack uses the above idea, inserting items $x$ into the shadow HLL sketch but resetting it as soon as the cardinality estimate increases from $B$ to $B+1$ (indicating that a non-target bucket has been hit by item $x$). Acting in this way precisely determines whether each inserted item hits the target buckets or not, so has no ``false positives'' which need to be eliminated through reprocessing of the list ${\cal L}$. This means that the attack can be carried out in a streaming fashion in a single pass over ${\cal X}$. Assuming $h$ behaves like a random function, we must perform a reset with probability $\frac{m-B}{m} = 1 - \frac{B}{m}$ per item inserted, and hence in total an expected number of $(1 - \frac{B}{m})\cdot 2^t$ resets.

>>>>

%\IEEEtitleabstractindextext{%
\begin{abstract}
Computing the count of distinct elements in massive data sets is a common task but naive approaches are memory-expensive. The HyperLogLog (HLL) algorithm (Flajolet \emph{et al.}, 2007) estimates a data set's cardinality while using significantly less memory than a naive approach at the cost of some accuracy. This trade-off makes the HLL algorithm very attractive for a wide range of applications such as database management and network monitoring, where an exact count may not be needed. Recently, the HLL algorithm has started to be proposed for use in scenarios where the inputs may be adversarially generated, for example detection of network scanning attacks. This prompts an examination of the performance of the HLL algorithm in the face of adversarial inputs. We show that in such a setting, the HLL algorithm's estimate of cardinality can be exponentially bad: when an adversary has access to the internals of the HLL algorithm and has some flexibility in choosing what inputs will be recorded, it can manipulate the cardinality estimate to be exponentially smaller than the true cardinality.
\end{abstract}
%}

@@

\begin{figure}[ht]
\centering
     \makebox[\linewidth]{
\fbox{
\begin{minipage}{40em}
\begin{flushleft}
\textit{Let $h : \mathcal{D} \rightarrow \{0, 1\}^{32}$ hash data from some domain $\mathcal{D}$ to binary $32$-bit strings.\\
Let $\rho(s)$ be the position of the leftmost $1$-bit of $s$: e.g., $\rho(1...) = 1, \rho(0001...) = 4, \rho(0^K) = K + 1.$\\}
\textbf{define} $m = 2^n$ with $n \in [4..16].$\\
\textbf{define} $\alpha_{16} = 0.673$; $\alpha_{32} = 0.697$; $\alpha_{64} = 0.709$; $\alpha_{m} = 0.7213/(1+ 1.079/m)$ for $m \ge 128$;\\
\vspace{0.5em}
\textbf{Algorithm} HyperLogLog (\textbf{input} $\mathcal{M}$ : multiset of items from domain $\mathcal{D}$).\\
\textbf{initialize} a collection of $m$ registers, $M[1], ..., M[m]$, to $0$;\\
\vspace{0.5em}
\textbf{for} $v \in \mathcal{M}$ \textbf{do}\\
    \hspace{2em} \textbf{set} $x := h(v);$\\
    \hspace{2em} \textbf{set} $j = 1 + <x_1x_2...x_{n}>_2$; \hfill \textit{\{the bucket index determined by the first $n$ bits of $x$\}}\\
    \hspace{2em} \textbf{set} $w := x_{n+1}x_{n+2}...$;\\
   \hspace{2em} \textbf{set} $M[j] :=$ max$(M[j], \rho(w))$;\\
\vspace{0.2em}
\textbf{compute} $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$; \hfill \textit{\{the raw HyperLogLog estimate\}}\\
\vspace{0.2em}
\textbf{if} $E \le \frac{5}{2}m$ \textbf{then}\\
    \hspace{2em}  \textbf{let} $V$ be the number of registers equal to $0$;\\
    \hspace{2em}  \textbf{if} $V \neq 0$ \textbf{then set} $E^*:=  m\cdot \ln(m/V)$ \textbf{else set} $E^*:= E$; \hfill \textit{\{small range correction\}}\\
\textbf{if} $E \le \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em} \textbf{set} $E^*:= E$; \hfill \textit{\{intermediate range—no correction\}}\\
\textbf{if} $E > \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em}  \textbf{set} $E^*:= -2^{32}$ln$(1 - E/2^{32})$; \hfill \textit{\{large range correction\}}\\
\textbf{return} \textit{cardinality estimate $E^*$ with typical relative error} $\pm1.04/\sqrt{m}$.
\end{flushleft}
\end{minipage}
}}
\caption{The HyperLogLog Algorithm from~\cite{hll}}
\label{fig:hll}
\end{figure}

@@

Here the attack is somewhat more complex. We begin by explaining the core idea, then a simple version of our attack, then a more complex but performant version, and finally, we give a heuristic performance analysis of the attack.

@@

\paragraph{Simple Attack:}
The simplistic version of our attack uses the above idea, inserting items $x$ into the shadow HLL sketch but resetting it as soon as the cardinality estimate increases from $B$ to $B+1$ (indicating that a non-target bucket has been hit by item $x$). Acting in this way precisely determines whether each inserted item hits the target buckets or not, so has no ``false positives'' which need to be eliminated through reprocessing of the list ${\cal L}$. This means that the attack can be carried out in a streaming fashion in a single pass over ${\cal X}$. Assuming $h$ behaves like a random function, we must perform a reset with probability $\frac{m-B}{m} = 1 - \frac{B}{m}$ per item inserted, and hence in total an expected number of $(1 - \frac{B}{m})\cdot 2^t$ resets.
